Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-05 paper2.qmd
section:6; process:write

  To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. This extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.

@@

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key CJ experimental design features, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. Finally, @sec-discussion discusses the findings, explores avenues for future research, and detail the challenges for future researchers.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general formulation, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases). However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of some traits or account for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022]. Indeed, the presence of the so-called *misfit* texts may already signal these limitations, as these are texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018], possibly due to larger discriminal dispersions or because they are genuine outliers--meaning, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, the BTL model neither accounts for these anomalies nor provides tools for addressing them, apart from excluding the "problematic" texts from analysis.

@@

<!-- "weak" writers consistently produce texts of low quality in all aspects that determine the quality of a text, such as grammatical accuracy, organization, and language use. In such cases, judges typically agree on their inferior quality, leading to a clear disfavoring of these texts compared to others. This agreement ultimately imply narrow dispersions for the discriminal differences related to these texts and, by extension, narrow discriminal processes for them. However, as texts improve along the trait continuum, with certain aspects showing improvement and others reaming unaltered, these variations introduce uncertainty into the judgments. Now, since judges often rely on various intricate characteristics of the texts to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], these changes lead to greater discrepancies in their evaluations. This divergence ultimately imply broader discriminal differences and, by extension, more varied discriminal processes for the texts. -->

@@

<!-- 3. But does it hold?  -->
Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, experimental designs often adopt the assumption of no correlation between stimuli based on an old theoretical justification. Specifically, @Thurstone_1927b argued that stimuli could be treated as uncorrelated because judges' biases--arising from two opposing and equally weighted effects occurring during the pairwise comparisons--would cancel each other out. This idea was later formalized by @Andrich_1978, who provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.

@@

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These characteristics, regardless of their relevance, are unlikely to be equally weighted or to oppose one another across comparisons. As a result, they may exert a disproportionate influence on judges' perceptions, generating biases that persist rather than cancel out. For example, this could occur when a judge assessing the argumentative quality of a text may place disproportionate emphasis on handwriting clarity, thereby favoring neatly written texts despite their weaker arguments. Moreover, because the discriminal process of stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II]. Evidence from traditional marking contexts confirms a similar phenomenon [@Morin_et_al_2018], and additional studies [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a] support the presence of these biases in CJ--reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.

@@

<!-- 3. But what are the challenges when this assumption is violated? -->
Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues. Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023] and even biases [@Wu_et_al_2022]. These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli. This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$.

@@

Finally, as discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969; @Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates. The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].

<!-- These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b. -->

@@

@@

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests. *Identification analysis* determines whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023]. Here, *estimands* represent the specific quantities researchers aim to determine (i.e., a parameter) [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate (e.g., a statistical model), while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear throughout this document clarifies these concepts. In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to causal inference. Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 2. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the conceptual-population model, which evaluates the impact of different teaching methods on students' writing ability. This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{iahbjk})$ and several related variables. The subscripts $i$ and $h$ identify the students who authored the texts (i.e., the individuals). The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli). The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for experimental conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, pp. 366-376]. Thus, the indexing system supports comparisons between different texts written by the same student $(i = h;$ $a \neq b)$ and between texts written by distinct students $(i \neq h;$ where $a = b$ is permitted$)$, each compared once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$. However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(i = h;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K};$ where $n_{J}>1$ and $n_{K}\geq1)$, as such comparison lacks practical relevance within the CJ framework. Here, $n_{J}$ indicates the total number of judges, and $n_{K}$ denotes the number of repeated judgments each judge performs.

@@

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory). Additionally, the SCM incorporates a key design feature of CJ experiments: the judges' biases $(B_{kj})$. This extension builds on the arguments presented in @sec-theory-issue1b, contending that the discriminal difference becomes an observable outcome only through judges' perceptions. Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons. This model defines the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$, and the preliminary set of structural equations, $F = { f_{O}, f_{D} }$, which capture the non-parametric dependencies among these variables.

@@

<!-- 3. segway for next section -->
However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals. Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.

@@

<!-- 4. segway for next section -->
In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model. These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and clarify the role of the sample and comparison mechanisms in CJ experiments (see @sec-theory-theoretical_SC). However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a. Since this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.

@@

<!-- 9. Computational implementation -->
Finally, we use *Bayesian inference methods* to convert the statistical model [-@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data. Bayesian inference offers three main advantages in this context. First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999]. This feature is essential for our implementation, as the proposed model is indeed overparameterized. Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. We primarily use this capability to define the error distribution and set the scale of latent variables [@Depaoli_2014]. Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014]. This feature is particularly relevant in CJ experiments, as researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.

@@

<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:58:20 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:12:00 - 00:14:30 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->
<!-- recording: Sven and Tine 25.04.22; time: 00:00:00 - 00:13:15 -->
<!-- recording: Sven and Tine 25.05.21; time: 00:00:00 - 00:03:30 -->
<!-- recording: Sven and Tine 25.05.21; time: 00:32:55 - 00:33:50 -->

<!-- Follow the structure of the paper: -->
<!-- 1. Talk about Thurstone -->
<!-- 2. Talk about the issues with the model of Thurstone -->
<!-- 3. Talk about what you did -->
<!-- 4. Potential benefits and extras that you can do now -->
<!-- 5. Challenges -->
<!-- 6. Conclusion: what we did and why is a great paper -->

@@

<!-- 2. Summary of the why the literature uses Case V, and its problems -->
Among these, Case V remains the most widely used model in empirical CJ research, mainly due to the widespread adoption of the BTL model. The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008]. Notably, although these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Moreover, because Thurstone's original goal was to produce a "coarse scaling" of  traits and allocate stimuli along this continuum [@Thurstone_1927a, pp. 269], his theory offered no guidance on how to use trait estimates for statistical inference. To address this gap, researchers in the CJ tradition have often separated trait estimation from hypothesis testing, relying on point estimates of the traits--such as BTL scores or their transformations--for inference. Notably, while this approach simplifies analysis, it can also introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods. The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.

The study then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur (see @sec-theory-issue1a and @sec-theory-issue1b) and when statistical inference is necessary (refer to @sec-theory-issue2). In particular, this model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical). By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.

@@

## Further research {#sec-discussion_RA}

Four current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.

>>>>

  To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. This extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons. It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.

@@

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ assessments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. Finally, @sec-discussion discusses the findings, explores avenues for future research, and detail the challenges for future researchers.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general formulation, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases). However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail accommodate for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022]. Indeed, the presence of the so-called *misfit texts*--texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]--may signal this limitation, as such discrepancies can arise from larger discriminal dispersions due to stimulus heterogeneity or because the texts are genuine outliers--that is, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, the BTL model's assumptions prevent it from adequately accounting for or addressing these anomalies, leaving exclusion of the "problematic" texts as the primary remedy.

@@

<!-- "weak" writers consistently produce texts of low quality in all aspects that determine the quality of a text, such as grammatical accuracy, organization, and language use. In such cases, judges typically agree on their inferior quality, leading to a clear disfavoring of these texts compared to others. This agreement ultimately imply narrow dispersions for the discriminal differences related to these texts and, by extension, narrow discriminal processes for them. However, as texts improve along the trait continuum, with certain aspects showing improvement and others reaming unaltered, these variations introduce uncertainty into the judgments. Now, since judges often rely on various intricate characteristics of the texts to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], these changes lead to greater discrepancies in their evaluations. This divergence ultimately imply broader discriminal differences and, by extension, more varied discriminal processes for the texts. -->

@@

<!-- 3. But does it hold?  -->
Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, assessment designs often adopt the assumption of no correlation between stimuli based on an old theoretical justification. Specifically, @Thurstone_1927b argued that stimuli could be treated as uncorrelated because judges' biases--arising from two opposing and equally weighted effects occurring during the pairwise comparisons--would cancel each other out. This idea was later formalized by @Andrich_1978, who provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.

@@

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate complex, multidimensional traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022]. Regardless of their relevance, these characteristics may not receive equal weight or consistently oppose one another across comparisons. As a result, they may exert a disproportionate influence on judges' perceptions, generating biases that persist rather than cancel out. For example, this could occur when a judge assessing the argumentative quality of a text may place disproportionate emphasis on handwriting clarity, thereby favoring neatly written texts despite their weaker arguments. Moreover, because the discriminal process of stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II]. While direct evidence for this exact scenario is limited, existing studies document the presence of judge bias in CJ [@Pollitt_et_al_2003, @vanDaal_et_al_2016, @Bartholomew_et_al_2020a], reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.

@@

<!-- 3. But what are the challenges when this assumption is violated? -->
Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues. Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023] and even biases [@Wu_et_al_2022]. This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$. Moreover, these inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli--potentially causing the model to incorrectly classify stimuli or judges as *misfits*.

@@

Finally, similar to what we discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and may appear as outliers under the BTL model [@Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates. The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].

@@

<!-- Under the Bradley–Terry model, the object residual ri asymptotically follows a standard normal distribution, and is independent of \beta_i. Note that the asymptotic distribution of Rij holds when Vij grows large, so it is important to exclude residuals show Rij for which Vij = 0.
If the absolute value of ri is very large, object i is regarded as an outlier. By definition, this happens if object i unexpectedly wins over a popular object j (i.e., \beta_j is positive) and Rij is large and positive, or if object i unexpectedly loses to an unpopular object j (\beta_j negative) and Rij is large and negative.
Too many outlying objects may imply that the model suffers from a lack of fit. As Liner and Amin (2004) pointed out, this is more likely to happen if some of the objects are involved in only a few comparisons with most of the others. [@Wu_et_al_2022, pp. 466] -->

@@

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ assessments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests. *Identification analysis* determines whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023]. Here, *estimands* represent the specific quantities researchers aim to determine (i.e., a parameter) [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate (e.g., a statistical model), while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear throughout this document clarifies these concepts. In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ assessment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to causal inference. Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key CJ assessment design features, such as the selection of judges, stimuli, and comparisons. In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ assessments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 2. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the conceptual-population model, which evaluates the impact of different teaching methods on students' writing ability. This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{iahbjk})$ and several related variables. The subscripts $i$ and $h$ identify the students who authored the texts (i.e., the individuals). The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli). The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for assessment conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, pp. 366-376]. Thus, the indexing system supports comparisons between different texts written by the same student $(i = h;$ $a \neq b)$ and between texts written by distinct students $(i \neq h;$ where $a = b$ is permitted$)$, each compared once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$. However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(i = h;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K};$ where $n_{J}>1$ and $n_{K}\geq1)$, as such comparison lacks practical relevance within the CJ framework. Here, $n_{J}$ indicates the total number of judges, and $n_{K}$ denotes the number of repeated judgments each judge performs.

@@

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory). Additionally, the SCM incorporates a key CJ design feature: the judges' biases $(B_{kj})$. This extension builds on the arguments presented in @sec-theory-issue1b, contending that the discriminal difference becomes an observable outcome only through judges' perceptions. Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons. This model defines the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$, and the preliminary set of structural equations, $F = { f_{O}, f_{D} }$, which capture the non-parametric dependencies among these variables.

@@

<!-- 3. segway for next section -->
However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ assessments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals. Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.

@@

<!-- 4. segway for next section -->
In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model. These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and even clarify the role of the sample and comparison mechanisms in CJ assessments (see @sec-theory-theoretical_SC). However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a. Since this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.

@@

<!-- 9. Computational implementation -->
Finally, we use *Bayesian inference methods* to convert the statistical model [-@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data. Bayesian inference offers three main advantages in this context. First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999]. This feature is essential for our implementation, as the proposed model is indeed overparameterized. Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. We primarily use this capability to define the error distribution and set the scale of latent variables [@Depaoli_2014]. Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014]. This feature is particularly relevant in CJ assessments, as researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.

@@

<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:58:20 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:12:00 - 00:14:30 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->
<!-- recording: Sven and Tine 25.04.22; time: 00:00:00 - 00:13:15 -->
<!-- recording: Sven and Tine 25.05.21; time: 00:00:00 - 00:03:30 -->
<!-- recording: Sven and Tine 25.05.21; time: 00:32:55 - 00:33:50 -->
<!-- recording: Sven and Tine 25.06.04; time: 00:00:00 - 00:05:40 -->

@@

<!-- 2. Summary of the why the literature uses Case V, and its problems -->
Among these, Case V remains the most widely used model in empirical CJ research, mainly due to the widespread adoption of the BTL model. The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008]. Although this substitution has minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], the simplifying assumptions of the BTL model--and by extension, of Case V--may fail to capture the complexity of some traits or account for heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Moreover, because Thurstone's original goal was to produce a "coarse scaling" of traits and allocate stimuli along this continuum [@Thurstone_1927a, pp. 269], his theory offered no guidance on how to use trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing, relying on point estimates--such as BTL scores or their transformations--for inference. Notably, while this approach simplifies analysis, it can also introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
Thus, to address the limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods. The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.

The approach then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur (see @sec-theory-issue1a and @sec-theory-issue1b) and when statistical inference is necessary (refer to @sec-theory-issue2). In particular, this model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical). By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement in CJ [@Perron_et_al_2015], while also improving the accuracy of statistical inferences.

@@

## Future research directions based on this study's approach {#sec-discussion_RA}
<!-- recording: Sven and Tine 25.06.04; time: 00:05:40 - 00:07:25 -->

Among the many potential directions for future research--both current and new--three avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates, as well as on the accuracy of statistical inferences. The following sections outline these avenues and explain how our approach provides a rigorous research framework to investigate them.
